%==============================================================================
%  Building GPT-2 from Scratch — CS224N Default Final Project
%  CENG-534, Izmir Institute of Technology (IZTECH)
%
%  Self-contained: compiles with a standard TeX Live / Overleaf pdfLaTeX
%  toolchain (no acl.sty or external style files required).
%  Build:  pdflatex main && bibtex main && pdflatex main && pdflatex main
%          (or simply: latexmk -pdf main.tex)
%==============================================================================
\documentclass[10pt,twocolumn]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{microtype}
\usepackage[margin=0.68in]{geometry}
\usepackage{times}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{multirow}
\usepackage[hidelinks]{hyperref}
\usepackage{listings}

\captionsetup{font=small,labelfont=bf}
\setlength{\columnsep}{0.28in}
\setlist{nosep,leftmargin=1.25em}

% Compact section spacing for a tidy 6--8 page layout.
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{1.0ex plus .2ex}{0.6ex}
\titlespacing*{\subsection}{0pt}{0.8ex plus .2ex}{0.4ex}

% Allow more floats per page so the many small tables pack tightly
% instead of spilling onto a near-empty final page.
\renewcommand{\topfraction}{0.92}
\renewcommand{\bottomfraction}{0.7}
\renewcommand{\textfraction}{0.06}
\renewcommand{\floatpagefraction}{0.75}
\setcounter{topnumber}{4}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{6}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  framesep=4pt,
  xleftmargin=2pt,
  aboveskip=0.6em,belowskip=0.4em,
}

\newcommand{\best}[1]{\textbf{#1}}

%==============================================================================
\title{\vspace{-2.0em}\textbf{Building GPT-2 from Scratch: Cloze-Style Paraphrase
Detection, Sonnet Generation, and a Stylistic Paraphrase Extension
(Modern\,$\rightarrow$\,Shakespearean)}}

\author{
  Hasan Semih Sel\c{c}uk \quad\quad Matin Huseynzade \\
  \small Department of Computer Engineering, Izmir Institute of Technology (IZTECH) \\
  \small CENG-534, Default Final Project (adapted from Stanford CS224N) \\
  \small \texttt{hasanselcuk@iyte.edu.tr}, \texttt{matinhuseynzade@iyte.edu.tr}
}
\date{}

%==============================================================================
\begin{document}
\maketitle

\begin{abstract}
\noindent
We implement the GPT-2 architecture from scratch (masked multi-head
self-attention, the pre-LayerNorm transformer block, and a decoupled-weight-decay
AdamW optimizer) and verify it against the course unit tests. Starting from
pretrained weights, we fine-tune the model on three downstream tasks. For
sentiment classification we reach a dev accuracy of \best{0.967} on CFIMDB
(full-model). For paraphrase detection we cast the Quora binary task as
cloze-style next-token prediction and obtain \best{0.889} dev accuracy. For
sonnet generation we run a decoding-temperature ablation and reach \best{41.49}
dev chrF at temperature $1.0$. As an original extension we build a
\emph{stylistic paraphrase generator} that rewrites modern English into
Shakespearean English; with source-side loss masking it improves from a
zero-shot baseline of $13.85$ to \best{35.83} chrF ($+21.97$). Beyond raw
numbers we contribute two analyses: (i) a prompt byte-pair-encoding ``footgun''
in which a trailing space in the prompt costs $6.3$ chrF, and (ii) a memory
recipe for stable variable-length full fine-tuning on the Apple Silicon MPS
backend, where we trained without any CUDA GPU.
\end{abstract}

%==============================================================================
\section{Introduction}
Large pretrained language models such as GPT-2 \cite{radford2019language} have
become the default backbone for both language understanding and generation. The
goal of this project is twofold: first, to \emph{understand} GPT-2 by building it
from its primitive components rather than calling a library; and second, to
\emph{adapt} it, through fine-tuning, to three downstream tasks of increasing
generative difficulty (sentiment classification, paraphrase detection, and
sonnet generation), and finally to a novel extension of our own.

A recurring theme of the project is the distinction between \emph{detecting} a
property of text (classification) and \emph{producing} text with a target
property (conditional generation). Paraphrase detection sits on the
classification side, while our extension (rewriting modern English into Shakespearean
style) sits squarely on the generation side and re-uses the same autoregressive
machinery as sonnet generation. This contrast motivated the design of the
extension.

\paragraph{Contributions.}
\begin{enumerate}
\item A complete, unit-test-verified implementation of GPT-2: scaled
dot-product masked attention, the pre-LayerNorm transformer block, and AdamW
with decoupled weight decay (\S\ref{sec:gpt2}).
\item A consistent cloze-prompt formulation for paraphrase detection together
with a sequence-length / left-truncation analysis that we found necessary for
both accuracy and memory (\S\ref{sec:paraphrase}).
\item A decoding-temperature ablation for sonnet generation
(\S\ref{sec:sonnet}, Table~\ref{tab:sonnet}).
\item \textbf{Extension:} stylistic paraphrase generation built on new parallel
data with source-side loss masking, reusing the sonnet model with \emph{no new
model class} (\S\ref{sec:extension}).
\item \textbf{Two analysis findings:} a prompt whitespace-tokenization footgun
worth $+6.3$ chrF, and a practical memory recipe for full fine-tuning on Apple
Silicon MPS (\S\ref{sec:analysis}).
\end{enumerate}

%==============================================================================
\section{Related Work}
GPT-2 \cite{radford2019language} is a decoder-only transformer
\cite{vaswani2017attention} trained with a causal language-modeling objective;
we re-implement its forward pass and train it with AdamW
\cite{loshchilov2019decoupled}. Casting classification as a fill-in-the-blank
(cloze) language-modeling problem follows the prompt-as-task line of work, most
directly PET \cite{schick2021exploiting} and the few-shot framing of GPT-3
\cite{brown2020language}; our paraphrase head reduces a binary decision to the
model's preference between the tokens \texttt{yes} and \texttt{no}. For open-ended
generation we use nucleus (top-$p$) sampling \cite{holtzman2020curious} to avoid
the degeneration that greedy and low-temperature decoding induce, and evaluate
generation with the character-$n$-gram metric chrF \cite{popovic2015chrf}, which
is well suited to the archaic morphology of our targets. Our extension builds on
the parallel ``No Fear Shakespeare'' corpus of Jhamtani et
al.~\cite{jhamtani2017shakespearizing}. Where they use a copy-enriched
sequence-to-sequence model, we instead fine-tune a single pretrained
decoder, trading task-specific architecture for transfer from pretraining.
Because we observed instability in multi-epoch full fine-tuning on our hardware,
parameter-efficient methods such as LoRA \cite{hu2022lora} are a natural avenue
for future work. The sentiment datasets are SST \cite{socher2013recursive} and
CFIMDB.

%==============================================================================
\section{Approach}
\subsection{GPT-2 implementation}
\label{sec:gpt2}
We implement the model bottom-up. \textbf{Masked multi-head self-attention}
computes, per head,
\begin{equation}
\mathrm{Attention}(Q,K,V)=\mathrm{softmax}\!\left(\frac{QK^{\top}}{\sqrt{d_k}}+M\right)V,
\end{equation}
where $M$ is a causal mask ($M_{ij}=-\infty$ for $j>i$) that forbids attending to
future positions, and the $h{=}12$ heads are concatenated and projected. Each
\textbf{transformer block} applies attention and a GELU MLP in the
\emph{pre-LayerNorm} configuration with residual connections,
$x \leftarrow x + \mathrm{Attn}(\mathrm{LN}(x))$ and
$x \leftarrow x + \mathrm{MLP}(\mathrm{LN}(x))$. The \textbf{full model} sums token
and positional embeddings, stacks $L{=}12$ blocks, applies a final LayerNorm, and
projects to the vocabulary with weights tied to the input embedding.

Our \textbf{AdamW} step implements bias-corrected first/second moments with
\emph{decoupled} weight decay applied directly to the parameters
\cite{loshchilov2019decoupled}:
\begin{align}
\hat m_t &= \tfrac{m_t}{1-\beta_1^t}, \quad \hat v_t = \tfrac{v_t}{1-\beta_2^t},\\
\theta_t &= \theta_{t-1} - \eta\Big(\tfrac{\hat m_t}{\sqrt{\hat v_t}+\epsilon}
+ \lambda\,\theta_{t-1}\Big).
\end{align}
We write the textbook bias-correction form for clarity; our code uses the
mathematically equivalent \emph{efficient} update of
\cite{loshchilov2019decoupled} (folding the correction into the effective step
size $\alpha_t=\eta\,\sqrt{1-\beta_2^t}/(1-\beta_1^t)$ and applying weight decay
with the original learning rate), which is exactly what the reference unit test
verifies.
Both \texttt{optimizer\_test.py} and \texttt{sanity\_check.py} pass
(\emph{``Your GPT2 implementation is correct!''}), confirming the forward pass and
the optimizer numerically match the reference.

\subsection{Sentiment classification}
We predict a sentiment label from the hidden state of the \emph{last} token of
the sequence, passed through dropout and a linear head. We compare two regimes:
\textbf{last-linear-layer} (the GPT-2 body is frozen and only the head is
trained) and \textbf{full-model} fine-tuning (all parameters updated). We
evaluate on SST-5 (5 classes) \cite{socher2013recursive} and CFIMDB (binary).

\subsection{Cloze-style paraphrase detection}
\label{sec:paraphrase}
We reduce Quora binary paraphrase classification to next-token prediction. A
shared template (\texttt{make\_cloze\_prompt}) ends with a position at which the
model must emit either \texttt{yes} (BPE id $8505$) or \texttt{no} (id $3919$);
the predicted label is whichever token receives higher probability.

\paragraph{Design decision: template consistency.} Training and evaluation use
the \emph{identical} cloze template. An earlier mismatch made dev look strong
while the held-out test score dropped. Consistency between train- and
test-time prompts proved essential.

\paragraph{Sequence length and left truncation.} We cap inputs at
$\texttt{max\_length}{=}128$ with \texttt{truncation\_side='left'}. Left
truncation is critical: it preserves the \emph{end} of the cloze prompt (where
the answer is produced) instead of discarding it. This helped both accuracy and
peak memory (\S\ref{sec:analysis}).

\subsection{Sonnet generation}
\label{sec:sonnet}
We fine-tune the model for autoregressive language modeling on Shakespeare's
sonnets. At test time the model is conditioned on the first three lines of a
sonnet and generates the remainder with nucleus (top-$p$) sampling
\cite{holtzman2020curious} at temperature $\tau$. We evaluate the generated
continuations against held-out gold with chrF \cite{popovic2015chrf} and ablate
$\tau$.

\subsection{Extension: stylistic paraphrase generation}
\label{sec:extension}
\paragraph{Task.} Rewrite a modern English sentence into Shakespearean English.
Unlike paraphrase \emph{detection}, this is conditional \emph{generation}. We
deliberately reuse the \texttt{SonnetGPT} infrastructure (teacher forcing,
top-$p$ generation, weight-tied projection), so that \textbf{no new model class is
introduced}, to isolate the effect of data and training objective.

\paragraph{Data.} We use the parallel ``No Fear Shakespeare'' corpus
\cite{jhamtani2017shakespearizing} (\texttt{harsh19/Shakespearizing-Modern-English}),
downloaded directly with \texttt{urllib}: \textbf{18{,}395 / 1{,}218 / 1{,}462}
aligned train/valid/test pairs after a light de-tokenization and an alignment
assertion.

\paragraph{Prompt and source-side loss masking.} Each example is formatted as
\texttt{Modern: \{m\}$\backslash$nShakespearean: \{s\}<eos>}. Crucially, the loss
is computed \emph{only over the target (Shakespearean) tokens}: prompt and
source tokens are masked with \texttt{ignore\_index = -100}, with the prompt
length computed per example. Concretely:
\begin{lstlisting}[language=Python]
logits = model(x)[:, :-1]
labels = x[:, 1:]
labels[:, :prompt_len-1] = -100  # mask source
labels[pad_positions]   = -100   # mask padding
loss = cross_entropy(logits, labels,
                     ignore_index=-100)
\end{lstlisting}
This focuses the gradient on \emph{producing} the style rather than on
reconstructing the (given) source sentence.

%==============================================================================
\section{Experiments}
\subsection{Data}
\textbf{Sentiment.} SST-5 ($8{,}544$ train sentences) and CFIMDB ($1{,}707$
train documents), using the course-provided splits. \textbf{Paraphrase.} The
Quora paraphrase CSVs shipped in our \texttt{data/} directory, containing
$283{,}011$ train / $40{,}430$ dev / $80{,}860$ test question
pairs.\footnote{\label{fn:quora}The handout (\S7.2) quotes a smaller subset
($141{,}506 / 20{,}215 / 40{,}431$); the CSVs distributed with our starter code
contain almost exactly twice as many \emph{unique} pairs. We verified this is not
a duplication artifact (the dev file has $40{,}430$ distinct ids and $40{,}430$
distinct, order-insensitive sentence pairs) and use the provided files in full;
all dev accuracies are measured on this dev file ($n{=}40{,}429$ labelled).}
\textbf{Sonnet.} $143$ training sonnets and a $12$-sonnet held-out \emph{test}
split (handout \S7.2). The starter data additionally ships a held-out
\emph{dev} set with gold (\texttt{TRUE\_sonnets\_held\_out\_dev.txt}), which we
use for temperature tuning; the official 12-sonnet test set is held out with
hidden labels. \textbf{Extension.} Jhamtani parallel corpus,
$18{,}395 / 1{,}218 / 1{,}462$. Metrics: accuracy for classification, chrF for
generation.

\subsection{Setup and hyperparameters}
\paragraph{Hardware.} All experiments ran on an \textbf{Apple M4 Pro (24\,GB
unified memory), with no CUDA GPU}, using the PyTorch \texttt{2.12.0} \textbf{MPS}
backend. Adapting variable-length full fine-tuning to MPS was a genuine
engineering constraint and is discussed in \S\ref{sec:analysis}. Table~\ref{tab:hp}
summarizes the per-task configuration; the optimizer is AdamW throughout
($\mathrm{lr}=10^{-5}$ for fine-tuning).

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Task} & \textbf{Configuration} \\
\midrule
Sentiment (last-lin.) & head only, batch 8 \\
Sentiment (CFIMDB full) & batch 4, max\_len 256, 5 ep. \\
Paraphrase & batch 16, max\_len 128 (L-trunc), 3 ep. \\
Sonnet & batch 8, 10 ep., top-$p$ 0.9 \\
Extension & batch 16, max\_len 128, 5 ep. \\
\bottomrule
\end{tabular}
\caption{Per-task training configuration (AdamW, lr $10^{-5}$ for fine-tuning;
``L-trunc'' = left truncation).}
\label{tab:hp}
\end{table}

%==============================================================================
\section{Results}
All accuracies and chrF scores below are our own measurements on the course
dev/held-out splits.

\begin{table*}[t]
\centering
\small
\begin{tabular}{@{}p{0.30\textwidth}p{0.33\textwidth}p{0.30\textwidth}@{}}
\toprule
\textbf{Modern (input)} & \textbf{Gold (reference)} & \textbf{Our model (generated)} \\
\midrule
I love rich Capulet's daughter.
& Then plainly know my heart's dear love is set / On the fair daughter of rich Capulet.
& I love rich Capulet's daughter. \\[0.4em]
Holy Saint Francis, this is a drastic change!
& Holy Saint Francis, what a change is here!
& Saint Francis, this is a sudden change! \\[0.4em]
Have you given up so quickly on Rosaline, whom you loved so much?
& Is Rosaline, whom thou didst love so dear, / So soon forsaken?
& Have you no more of Rosalind, whom thou loved so much? \\
\bottomrule
\end{tabular}
\caption{Qualitative stylistic-paraphrase examples (modern $\rightarrow$
Shakespearean, \S\ref{sec:analysis}). The model adopts archaic forms
(\emph{thou}, \emph{didst}) and a period register while preserving meaning.}
\label{tab:qual}
\end{table*}

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
& \multicolumn{2}{c}{\textbf{Last-linear}} & \multicolumn{2}{c}{\textbf{Full-model}} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
\textbf{Dataset} & acc & F1 & acc & F1 \\
\midrule
SST-5  & 0.461 & 0.366 & \best{0.513} & \best{0.467} \\
CFIMDB & 0.857 & 0.857 & \best{0.967} & \best{0.967} \\
\bottomrule
\end{tabular}
\caption{Sentiment classification, dev accuracy and macro-F1. The handout dev
baselines are $0.462 / 0.513 / 0.861 / 0.976$ (SST last / SST full / CFIMDB last /
CFIMDB full). We match SST full exactly ($0.513$) and SST last to within rounding
($0.461$ vs.\ $0.462$); CFIMDB last is marginally lower ($0.857$ vs.\ $0.861$).
Since we report a single run, this small gap is within expected run-to-run
variation and is not directly comparable. The one real gap is CFIMDB full
($0.967$ vs.\ $0.976$); we initially hypothesised this stemmed from the reduced
batch size ($4$) we needed to fit MPS memory, but a controlled experiment
(\S\ref{sec:analysis}) does \emph{not} support that explanation.}
\label{tab:sentiment}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Model} & \textbf{Epoch 0} & \textbf{Epoch 1} & \textbf{Epoch 2} \\
\midrule
GPT-2 small, full FT, cloze & 0.863 & 0.879 & \best{0.889} \\
\bottomrule
\end{tabular}
\caption{Paraphrase detection (Quora, dev accuracy). Accuracy rises
monotonically over three epochs of full fine-tuning with the consistent cloze
template and left truncation.}
\label{tab:paraphrase}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{Temperature $\tau$} & 0.7 & 0.9 & 1.0 & 1.2 \\
\midrule
dev chrF & 39.01 & 40.63 & \best{41.49} & 41.12 \\
\bottomrule
\end{tabular}
\caption{Sonnet generation, dev chrF (top-$p=0.9$ fixed). chrF improves
monotonically up to $\tau=1.0$; the $1.0$ vs.\ $1.2$ difference is only
$0.37$ chrF, within single-run noise, so we select $\tau=1.0$. Lower temperature
($0.7$) shows visibly more repetition.}
\label{tab:sonnet}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lc@{}}
\toprule
\textbf{Model} & \textbf{chrF} \\
\midrule
Zero-shot pretrained GPT-2 (no fine-tune) & 13.85 \\
Fine-tuned (source-side masking) & \best{35.83} \\
\midrule
$\Delta$ (gain) & \best{+21.97} \\
\bottomrule
\end{tabular}
\caption{Extension: stylistic paraphrase (test chrF, $n=1462$). Fine-tuning with
source-side loss masking more than doubles the zero-shot chrF.}
\label{tab:extension}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lc@{}}
\toprule
\textbf{Prompt design} & \textbf{chrF} \\
\midrule
\texttt{"Shakespearean: "} (trailing space) & 29.52 \\
\texttt{"Shakespearean:"} + \texttt{" \{target\}"} & \best{35.83} \\
\bottomrule
\end{tabular}
\caption{Extension ablation: prompt design (test chrF). Removing the trailing
space from the prompt tag, and instead prefixing the space to the target, is
worth $+6.3$ chrF (see \S\ref{sec:analysis}).}
\label{tab:promptablation}
\end{table}

\paragraph{Sentiment (Table~\ref{tab:sentiment}).} Full-model fine-tuning beats
the frozen-body head on both datasets, dramatically so on the binary CFIMDB task
($0.857 \rightarrow 0.967$). On the harder 5-class SST the absolute numbers are
lower ($0.461 \rightarrow 0.513$), as expected for fine-grained sentiment.

\paragraph{Paraphrase (Table~\ref{tab:paraphrase}).} The cloze formulation
improves steadily to $0.889$ dev accuracy at epoch~2. We stress that this number
is measured on our larger dev split (about twice the handout subset; see
footnote~\ref{fn:quora}), so it is \emph{not} directly comparable to handout
baselines computed on the smaller subset.

\paragraph{Sonnet (Table~\ref{tab:sonnet}).} Sweeping decoding temperature shows
chrF rising monotonically to $\tau=1.0$ (chrF $41.49$). The gap to $\tau=1.2$
($41.12$) is only $0.37$ chrF, within single-run noise, so we select $\tau=1.0$
for the test submission rather than claiming a sharp optimum. As a second,
non-chrF axis, fine-tuning lowers held-out dev \emph{perplexity} from $99.1$
(zero-shot pretrained GPT-2) to $59.0$ (lower is better). We report this for
reproducibility and to show fine-tuning helps on a second metric, not as a
standalone claim of absolute quality.

\paragraph{Extension (Tables~\ref{tab:extension}--\ref{tab:promptablation}).}
Fine-tuning lifts chrF from $13.85$ to $35.83$ ($+21.97$). Both the zero-shot
baseline and the fine-tuned model are evaluated with the \emph{corrected}
(no-trailing-space) prompt and identical decoding, so this gain is not confounded
by the whitespace effect studied in Table~\ref{tab:promptablation}. A single
prompt-design change accounts for $6.3$ of the fine-tuned points.

%==============================================================================
\section{Analysis}
\label{sec:analysis}

\paragraph{Qualitative: stylistic paraphrase.}
Table~\ref{tab:qual} shows representative model outputs against gold. The model
reliably injects archaic morphology (\emph{hath}, \emph{doth}, \emph{thou}) and
re-orders clauses toward an early-modern register, while staying faithful to the
source meaning. Failures are dominated by occasional sampling-induced repetition
(e.g.\ ``pray you, pray you'') rather than semantic drift.

\paragraph{The prompt whitespace footgun ($+6.3$ chrF).}
Our single most instructive finding (Table~\ref{tab:promptablation}). GPT-2's
byte-pair encoding attaches a leading space to the \emph{following} token
(``\texttt{ Shakespearean}'' is a different token than ``\texttt{Shakespearean}'').
When the prompt tag \emph{ended} with a space (\texttt{"Shakespearean: "}), the
model's first decoding step had to emit a token with no leading space, an
out-of-distribution situation, and it frequently produced junk artifacts
(\texttt{\~{}\~{}\~{}\~{}}, \texttt{----}, \texttt{\_\_\_\_}) before recovering.
Moving the space off the prompt and onto the target (tag
\texttt{"Shakespearean:"}, target \texttt{" \{s\}"}) aligns generation with the
BPE convention and recovers $6.3$ chrF. This is a subtle, reproducible tokenizer
pitfall that generalizes to any prompted GPT-2 generation.

\paragraph{Decoding behaviour.}
The temperature sweep (Table~\ref{tab:sonnet}) traces the
degeneration/divergence trade-off of \cite{holtzman2020curious}: at $\tau=0.7$ the
sonnet model shows visibly more repetition (e.g.\ ``this, this\ldots''),
consistent with the lower chrF; at $\tau=1.2$ it begins to wander; $\tau=1.0$
balances diversity and fidelity. We quantify this with corpus-level diversity of
the dev generations (pooled $n$-grams over the $12$ dev sonnets, fixed seed and
conditioning): the fraction of repeated $4$-grams falls from $0.17$ at
$\tau{=}0.7$ to $0.008$ at $\tau{=}1.0$ and $0.002$ at $\tau{=}1.2$, while
distinct-$2$ rises from $0.51$ to $0.81$ to $0.88$. The low-temperature
repetition is therefore real and monotone, not a cherry-picked example; these
remain single-run observations. There is also a methodological tension worth
stating: chrF is an $n$-gram-overlap metric, the
very family \cite{holtzman2020curious} argue underweights degeneration, so we
read the temperature trend together with the qualitative samples rather than
relying on chrF alone. We use chrF because it is the course's grading metric.

\paragraph{Systems finding: full fine-tuning on Apple Silicon (MPS).}
With no CUDA GPU available, we trained entirely on the MPS backend, where
variable-length batches fragment the memory allocator: because each batch has a
different shape, the cached pool grows until it triggers an out-of-memory error
(we observed peaks above $25$\,GiB on CFIMDB and a crash mid-epoch on the
$22$k-batch Quora run). Our recipe for stability was: (i) \emph{bound the
sequence length} and use small batches; (ii) call
\texttt{torch.mps.empty\_cache()} periodically (every 10--50 steps); and (iii)
wrap all evaluation in \texttt{torch.no\_grad()}. With these in place, CFIMDB
full fine-tuning and the 3-epoch paraphrase run completed without OOM. We report
this as a practical recipe we found necessary for GPU-free full fine-tuning, not
as a novel method.

\paragraph{Does batch size explain the CFIMDB gap?}
Table~\ref{tab:sentiment} noted that CFIMDB full ($0.967$) trails the handout
($0.976$), and we hypothesised the small batch ($4$, forced by MPS memory) was the
cause. We tested this with a controlled gradient-accumulation experiment: holding
seed, learning rate, epochs ($5$), dropout, and micro-batch fixed, we raised the
\emph{effective} batch from $4$ to $32$ by accumulating over $8$ micro-batches. The
accumulation-step-$1$ control reproduced our original score exactly ($0.9673$),
confirming the implementation. The per-epoch dev accuracies tell the story: effective
batch $4$ gives $0.967, 0.967, 0.947, 0.927, 0.910$ (it peaks within one epoch and
then overfits), whereas effective batch $32$ gives $0.551, 0.653, 0.922, 0.947,
0.959$ (it climbs monotonically and is \emph{still improving} at the $5$-epoch budget,
not yet converged). So a larger batch does \emph{not} reach a higher score within the
same budget; the data therefore \emph{do not support} the hypothesis that the small
batch caused the gap. We are careful not to over-read this: the two final scores
differ by only $0.008$ on a single seed and the large-batch run had not converged, so
we cannot claim the larger batch is genuinely \emph{worse}, only that it fails to
close the gap. The residual gap to $0.976$ is unexplained and, under our single-seed
protocol, may reflect initialisation or other run-to-run variation. We consider this
informative: a plausible hypothesis, tested rather than asserted.

\paragraph{Limitations.}
We use the small GPT-2 ($124$M). The sonnet corpus is tiny ($143$ sonnets), so
the model overfits beyond $\sim$2 epochs (dev chrF degrades). Multi-epoch full
fine-tuning was unstable on MPS for the largest runs, which is precisely the
regime where parameter-efficient methods such as LoRA \cite{hu2022lora} would
help, both for memory and for overfitting.

%==============================================================================
\section{Conclusion and Future Work}
We built GPT-2 from its primitives and fine-tuned it across sentiment
classification (CFIMDB dev acc $0.967$), cloze-style paraphrase detection (Quora
dev acc $0.889$), and sonnet generation (dev chrF $41.49$), then extended it to a
stylistic paraphrase generator that improves from $13.85$ to $35.83$ chrF
($+21.97$). Two analyses, a $6.3$-chrF prompt tokenization footgun and a memory
recipe for MPS full fine-tuning, turned out to matter as much as the headline
numbers. Future work: parameter-efficient fine-tuning (LoRA \cite{hu2022lora}) to
enable a larger backbone (GPT-2 medium) and more stable multi-epoch training;
preference-based alignment (e.g.\ DPO) for sonnet quality; and a wider top-$p$ /
temperature search.

%==============================================================================
\appendix
\section{Reproducing the Results}
\label{app:repro}
Every number in Tables~\ref{tab:sentiment}--\ref{tab:promptablation} was computed
by us on the dev / held-out splits and can be regenerated as follows. All commands
run from the repository root with the course environment active
(\texttt{conda activate cs224n\_dfp};
\texttt{export PYTORCH\_ENABLE\_MPS\_FALLBACK=1}). The full procedure is also in
\texttt{REPRODUCE.md}.

\paragraph{Fast path (seconds, CPU, no retraining).} Re-score the saved
predictions in \texttt{predictions/} against the gold labels in \texttt{data/};
this reproduces Tables~\ref{tab:sentiment}--\ref{tab:extension} exactly:
\begin{lstlisting}[language=bash]
python reproduce_metrics.py
\end{lstlisting}
The script loads no model: it applies scikit-learn accuracy/F1 and the course
\texttt{evaluation.test\_sonnet} / sacrebleu \texttt{CHRF} scorers to the
committed artifacts, so the reported numbers are independently checkable.

\paragraph{Full re-training / generation.} The runs that produced those artifacts:
\begin{lstlisting}[language=bash]
# Part 0: implementation sanity checks
python optimizer_test.py && python sanity_check.py

# Table 3: sentiment (dev acc / macro-F1)
python classifier.py \
  --fine-tune-mode last-linear-layer --use_gpu --lr 1e-3
python classifier.py \
  --fine-tune-mode full-model --use_gpu --lr 1e-5

# Table 4: paraphrase (Quora dev acc)
python paraphrase_detection.py \
  --use_gpu --epochs 3 --batch_size 16 --lr 1e-5

# Table 5: sonnet (dev chrF temperature sweep)
python sonnet_generation.py --use_gpu --epochs 10 --lr 1e-5
for T in 0.7 0.9 1.0 1.2; do
  python eval_sonnet_dev.py --use_gpu \
    --temperature $T --top_p 0.9
done

# Tables 6-7: stylistic paraphrase extension
python prepare_style_data.py
python stylistic_paraphrase.py \
  --use_gpu --epochs 5 --batch_size 16 --lr 1e-5

# Sonnet diagnostics: held-out perplexity (99.1 zero-shot -> 59.0 FT) + per-temperature
# corpus-level distinct-n / rep-4 (no retraining; uses the saved checkpoint)
python analyze_sonnet.py --use_gpu

# CFIMDB batch-size hypothesis test (Analysis): effective batch 4 vs 32 via
# gradient accumulation, all else fixed -> 0.9673 vs 0.9592 (hypothesis not supported)
python run_cfimdb_batch.py all
\end{lstlisting}
Test-set scores are intentionally omitted: the official test labels are hidden,
so we report dev / held-out numbers throughout.

%==============================================================================
{\small
\bibliographystyle{plain}
\bibliography{references}
}

\end{document}
